---
title: "MDS plot and Venn diagram of wild sea otter-associated sample types"
author: "Natasha Dudek"
date: "03/12/2018"
output: html_document
---

In this script we will:
1. Plot an ordination of gingival, rectal, and otter-associated seawater samples (Figure 2C) and perform significance testing to determine whether the composition of sample types is distinct (Figure 2C)
2. Plot a Venn diagram of sets of gingival, rectal, and otter-associated seawater ASVs (Figure 2D)

Set paths + working directory and load phyloseq object (wild sea otter dataset)

```{r}
path <- file.path("/Users", "path", "to", "dir", fsep="/") # path to working dir
setwd(path)
getwd()
physeq <- readRDS(path, "physeq_lane1.rds")
```

Load libraries

```{r}
library("ggplot2"); packageVersion("ggplot2")
library("nVennR"); packageVersion("nVennR")
library("plyr"); packageVersion("plyr")
library("phyloseq"); packageVersion("phyloseq")
```

ggplot2 package theme set

```{r}
theme_set(theme_bw())
```

Filter out non-relevant samples   
- Select only wild sea otters (facility = CDFW)  
- Select only one sample per individual sea otter and sample type   
- Select only gingival, rectal, and seawater samples

```{r}
physeq <- subset_samples(physeq, facility == "CDFW") 
physeq <- subset_samples(physeq, duplicate_capture == "no")
physeq <- subset_samples(physeq, sample_type != "fecal")
```

How many of each sample type do we have?

```{r}
table(sample_data(physeq)$sample_type)
```

## Ordinate

Remove ASVs that do not appear more than 5 times in at least 1% of samples. This helps protect against ASVs with small mean & trivially large C.V.   

```{r}
temp = genefilter_sample(physeq, filterfun_sample(function(x) x > 5), A=0.01*nsamples(physeq))
physeq_filtered = prune_taxa(temp, physeq)
```

Perform MDS ordination using four diversity metrics  
These metrics were selected such that all combinations of the following characteristics are accounted for:  
a) Presence/absence only, b) relative abundance  
a) Phylogenetically uniformed, b) phylogenetically informed  

```{r}

dist_methods = c("jaccard","unifrac","bray","wunifrac")

plist <- vector("list", length(dist_methods))
names(plist) = dist_methods


for( i in dist_methods ){
    iDist <- distance(physeq_filtered, method=i)
    iMDS  <- ordinate(physeq_filtered, "MDS", distance=iDist)
 	evals <- iMDS$values$Eigenvalues
    p <- NULL
    p <- plot_ordination(physeq_filtered, iMDS, color="sample_type") +
  coord_fixed(sqrt(evals[2] / evals[1]))
    plist[[i]] = p
}

df = ldply(plist, function(x) x$data)
names(df)[1] <- "distance"
p = ggplot(df, aes(Axis.1, Axis.2, color=sample_type))
p = p + geom_point(size=1, alpha=0.7)
p = p + facet_wrap(~distance, scales="free")
p = p + scale_color_manual(values = c("#4fc657", "#ba7d28", "#269af2"))
p
```

Regardless of distance metric, samples of different types appear to be fairly distinct. For the main manuscript, print a figure using only the unifrac distance metric and scale axes to proportion of variance explained.

```{r}
physeq_filtered.ord <- ordinate(physeq_filtered, "MDS", "unifrac")

evals <- physeq_filtered.ord$values$Eigenvalues

p2 = plot_ordination(physeq_filtered, physeq_filtered.ord, color="sample_type") +
  coord_fixed(sqrt(evals[2] / evals[1]))
  
p2 = p2 + scale_color_manual(values = c("#4fc657", "#ba7d28", "#269af2"))
ggsave('Fig2C.pdf', plot = p2) 
p2
```

## Plot the Venn diagram showing sets of ASVs in gingival, rectal, and seawater samples

Use phyloseq object before filtering to remove ASVs that do not appear more than 5 times in at least 1% of samples. Instead we will just remove ASVs with zero reads.

```{r}
physeq = filter_taxa(physeq, function(x) sum(x) > 0, TRUE)
```

Select only sea otters for whom there is a gingival, rectal, and adjacent seawater sample  
i.e. whose BRD number shows up three times (remember there is a max of only one of each sample type per sea otter)

```{r}
brd_list <- sample_data(physeq)$brd_number
count_brds <- table(brd_list)
keeper_brds <- names(count_brds)[count_brds == 3]
venn_physeq <- subset_samples(physeq, brd_number %in% keeper_brds) 
```

How many sea otters have samples meeting these criteria?

```{r}
length(names(count_brds)[count_brds == 3])
```

Select only sea otters for which all three samples were collected on the same day

```{r}
day <- sample_data(venn_physeq)$day_captured
month <- sample_data(venn_physeq)$month_captured
year <- sample_data(venn_physeq)$year_captured

u_keeper_brds <- unique(keeper_brds)
strict_keepers <- c()
for (i in u_keeper_brds){
    temp_physeq <- subset_samples(venn_physeq, brd_number == i) 
    
    day <- sample_data(temp_physeq)$day_captured
    month <- sample_data(temp_physeq)$month_captured
    year <- sample_data(temp_physeq)$year_captured
    
    if ((!(NA %in% day)) & (!(NA %in% month)) & (!(NA %in% year)) & (length(unique(day)) == 1) & (length(unique(month)) == 1) & (length(unique(year)) == 1)){
        strict_keepers <- append(strict_keepers, i)   
    } 
}

venn_physeq <- subset_samples(physeq, brd_number %in% strict_keepers) 
```

How many sea otters are we looking at now?

```{r}
length(strict_keepers)
```

Rarify samples with replacement - need equal sampling depth  
Note: trimOTUs function removes all OTUs not present in the rarified dataset after subsampling

```{r}
min(sample_sums(venn_physeq))
rarified_physeq <- rarefy_even_depth(venn_physeq, sample.size = min(sample_sums(venn_physeq)), rngseed = 290518, replace = TRUE, trimOTUs = TRUE, verbose = TRUE)
```

Sampling depth

```{r}
unique(sample_sums(rarified_physeq))
```

Number of ASVs found in all samples

```{r}
total_num <- ntaxa(rarified_physeq)
total_num
```

Number of ASVs found in rectal samples ONLY

```{r}
oral_water <- subset_samples(rarified_physeq, sample_type == "oral" | sample_type == "water")
oral_water = filter_taxa(oral_water, function(x) sum(x) > 0, TRUE)
num_rectal_only <- total_num - ntaxa(oral_water)
num_rectal_only
```

Number of ASVs found in oral samples ONLY

```{r}
rectal_water <- subset_samples(rarified_physeq, sample_type == "rectal" | sample_type == "water")
rectal_water = filter_taxa(rectal_water, function(x) sum(x) > 0, TRUE)
num_oral_only <- total_num - ntaxa(rectal_water)
num_oral_only
```

Number of ASVs found in water samples ONLY

```{r}
oral_rectal <- subset_samples(rarified_physeq, sample_type == "rectal" | sample_type == "oral")
oral_rectal = filter_taxa(oral_rectal, function(x) sum(x) > 0, TRUE)
num_water_only <- total_num - ntaxa(oral_rectal)
num_water_only
```

Number of ASVs shared by rectal and oral samples only 

```{r}
water <- subset_samples(rarified_physeq, sample_type == "water")
water = filter_taxa(water, function(x) sum(x) > 0, TRUE)
num_water <- ntaxa(water)
num_shared_oral_rectal = total_num - num_water - num_oral_only - num_rectal_only 
num_shared_oral_rectal
```

Number of ASVs shared by oral and water samples only

```{r}
rectal <- subset_samples(rarified_physeq, sample_type == "rectal")
rectal = filter_taxa(rectal, function(x) sum(x) > 0, TRUE)
num_rectal <- ntaxa(rectal)
num_shared_oral_water = total_num - num_rectal - num_oral_only - num_water_only 
num_shared_oral_water
```

Number of ASVs found in rectal and water samples only 

```{r}
oral <- subset_samples(rarified_physeq, sample_type == "oral")
oral = filter_taxa(oral, function(x) sum(x) > 0, TRUE)
num_oral <- ntaxa(oral)
num_shared_rectal_water = total_num - num_oral - num_water_only - num_rectal_only 
num_shared_rectal_water
```

Number of ASVs found in all three sample types

```{r}
num_shared = total_num - num_oral_only - num_rectal_only - num_water_only - num_shared_oral_rectal - num_shared_oral_water - num_shared_rectal_water
num_shared
```

Now make the Venn diagram -- adjust labels later in Inkscape.

```{r}
myV3 <- createVennObj(nSets = 3, sSizes = c(rep(1, 7)))

myV3 <- setVennRegion(myV3, region = c("Group1"), value = as.numeric(num_oral_only[1]))
myV3 <- setVennRegion(myV3, region = c("Group2"), value = as.numeric(num_rectal_only[1]))
myV3 <- setVennRegion(myV3, region = c("Group3"), value = as.numeric(num_water_only[1]))

myV3 <- setVennRegion(myV3, region = c("Group1","Group2"), value = as.numeric(num_shared_oral_rectal[1]))
myV3 <- setVennRegion(myV3, region = c("Group1","Group3"), value = as.numeric(num_shared_oral_water[1]))
myV3 <- setVennRegion(myV3, region = c("Group2","Group3"), value = as.numeric(num_shared_rectal_water[1]))

myV3 <- setVennRegion(myV3, region = c("Group1","Group2","Group3"), as.numeric(num_shared[1]))

myV3 <- plotVenn(nVennObj = myV3)
showSVG(myV3, setColors = c("#4fc657", "#ba7d28", "#269af2"), opacity = 0.1, borderWidth = 3, fontScale = 2, outFile="Fig2D.svg")
```
